Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-28 numerical.tex
updating due to barker latter 2015

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield similar results.) This
is similar to strictly isothermal perturbations in a vertically
isothermal disk considered in \cite{nelson13} and \cite{mcnally14}.

@@

With rapid thermal relaxation, growth rates are expected to diminish
for both $\khat\to 0$ and $\khat\to \infty$. This is because for
$\khat\ll 1$ the radial wavelength is large, but such disturbances are 
stabilized by by epicyclic motions ($|\omega|\sim
\kappa$). Disturbances with $\khat\gg 1$ are radially localized, but  
there is no energy change when fluid elements are perturbed at fixed
cylindrical radius whilst conserving its angular momentum (applicable
to axisymmetric perturbations).

Fig. \ref{lowfreq_eigenfunc} plots the eigenfunctions $W$ and $\delta 
v_z$ for the fundamental mode with $\khat=10$ (i.e. the mode with
maximum growth rate in Fig. \ref{iso_eigen_kx}). We find $W\propto z$
throughout most of the disk, corresponding to a constant vertical
velocity perturbation, as expected for the fundamental mode (see
Appendix \ref{iso_explicit}). We show a meridional visualization of
this mode in Fig. \ref{lowfreq_eigenfunc_2d}. The $x$ axis is
stretched for clarity; radial velocities are in fact typically much 
smaller than vertical velocities. Most of the meridional kinetic energy is
contained within $\sim 2H$ of the midplane due to the density 
stratification.

@@

\subsubsection{Case study: $\khat=10$}
In Fig. \ref{compare_modes_iso_kx10} we plot eigenvalues of unstable 
modes with $\khat=10$. Black dots correspond to our fiducial setup. 
We also plot other calculations with different
vertical boundary conditions: rigid boundary and $\zmax=5H$ (red
crosses),  free surface and $\zmax = 3H$ (blue triangles), free surface
and $\zmax=7H$ (green crosses).

@@

The `tail' end of modes with decreasing $\nu$ to the right of
Fig. \ref{compare_modes_iso_kx10} is not predicted by our 
analysis. They have lower growth rates than modes consistent with our
analysis (i.e. those along the lines). These modes differ significantly between
cases and are thus likely associated with boundary conditions. % We do
                                % not consider them  
                                % further.

@@

\subsubsection{Comment on `surface modes'}\label{surf_comment}
Our fiducial setup with $\khat=10$ in 
Fig. \ref{compare_modes_iso_kx10} (black dots) differs to that from
\cite{nelson13} and \cite{mcnally14} in the lack of `surface modes'. 
These modes lie almost parallel to the vertical axis in mode 
diagrams such as Fig. \ref{compare_modes_iso_kx10}. We find them for 
sufficiently large $\zmax$ and/or $\khat$.

To illustrate this, we plot the mode diagram for $\khat=30$ in  
Fig. \ref{compare_modes_iso_kx30} with different boundary 
conditions. Surface modes are readily identified for 
$\zmax\geq 5H$ and are marked by `s'. They have the largest growth
rates but require sufficiently large vertical domain size and their
eigenvalues vary with $\zmax$.

Due to their dependence on boundary conditions, we do not consider
surface modes further in detail, which have
already been discussed in the aforementioned studies. In fact, we will
see in the next section that surface modes are more effectively
stabilized by finite thermal relaxation than the fundamental mode.

>>>>

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield similar results.) 
This setup mimics strictly isothermal perturbations in a vertically
isothermal disk, as considered by \cite{nelson13}, \cite{mcnally14} and
\cite{barker15}. %Note, however, that \citeauthor{nelson13} and
% \citeauthor{barker15} employed the  approximations described in \S\ref{analytical} throughout their analysis, which are relaxed in our
% numerical study.  
For completeness we review this case before considering the effect of
thermal relaxation.

%  In particular, we focus
% on results most relevant to our eventual application to protoplanetary disks.

% The reader is referred
% to the above works for a detailed look at different VSI modes. Here,
% we present selected results most relevant to our eventual application
% to protoplanetary disks.

@@

% With rapid thermal relaxation, growth rates are expected to diminish
% for both $\khat\to 0$ and $\khat\to \infty$. This is because for
% $\khat\ll 1$ the radial wavelength is large, but such disturbances are 
% stabilized by by epicyclic motions ($|\omega|\sim
% \kappa$). Disturbances with $\khat\gg 1$ are radially localized, but  
% there is no energy change when fluid elements are perturbed at fixed
% cylindrical radius whilst conserving its angular momentum (applicable
% to axisymmetric perturbations).

Fig. \ref{lowfreq_eigenfunc} plots the eigenfunctions $W$ and $\delta 
v_z$ for the fundamental mode with $\khat=10$ (which has the 
maximum growth rate in Fig. \ref{iso_eigen_kx}). We find $W\propto z$
throughout most of the disk, corresponding to a constant vertical
velocity perturbation, as expected for the fundamental mode (see
Appendix \ref{iso_explicit}). We show a meridional visualization of
this mode in Fig. \ref{lowfreq_eigenfunc_2d}. The $x$ axis is
stretched for clarity; radial velocities are in fact typically much 
smaller than vertical velocities. Most of the meridional kinetic energy is
contained within $\sim 2H$ of the midplane due to the density 
stratification.

@@

\subsubsection{Case study: $\khat=10$}
In Fig. \ref{compare_modes_iso_kx10} we plot eigenvalues of all unstable 
modes with $\khat=10$. Black dots correspond to our fiducial setup. 
We also plot other calculations with different
vertical boundary conditions: rigid boundary and $\zmax=5H$ (red
crosses),  free surface and $\zmax = 3H$ (blue triangles), free surface
and $\zmax=7H$ (green crosses).

@@

The `tail' end of modes with decreasing $\nu$ to the right of
Fig. \ref{compare_modes_iso_kx10} are in fact also unstable inertial modes, like those
along the analytic lines, but are strongly affected by the presence of 
the imposed vertical boundaries \citep{barker15}.

% They have lower growth rates than modes consistent with our
% analysis (i.e. those along the lines). These modes differ significantly between
% cases and are thus likely associated with boundary conditions. 
% We do
% not consider them  
% further.

@@

\subsubsection{Comment on `surface modes'}\label{surf_comment} 
Our fiducial setup with $\khat=10$ in 
Fig. \ref{compare_modes_iso_kx10} (black dots)  lack `surface modes'
found by \cite{nelson13}, \cite{mcnally14}  and 
by \cite{barker15}.  We illustrate this with the mode
diagram for $\khat=30$ in Fig. \ref{compare_modes_iso_kx30}. Surface
modes lie almost parallel to the vertical axis
in such diagrams. They are readily identified for  $\zmax\geq 5H$ and
are marked by `s'.  We typically find an increasing
number of them with increasing $\zmax$ and/or $\khat$,
consistent with \cite{barker15}.

As explained by \cite{barker15}, the occurence of surface modes in the
vertically isothermal disk is due to imposing a finite vertical domain
size. The fastest growing mode is typically a surface mode, if they
exist, but their association with vertical boundaries means that
$\mathrm{max}(\nu)$ depends on the choice of $\zmax$. Thus, the
vertically isothermal disk is not useful for studying surface modes. 
%Thus, seeking the most unstable mode in a vertically isothermal disk
%may not yield robust results because this

We will see in the next section that surface modes, as well
as other higher order modes, are more effectively stabilized by finite
thermal relaxation than the fundamental mode. Since we are primarily
interested in characterizing a maximum thermal timescale that allows
the VSI, we can focus on the fundamental mode. For a detailed analysis of 
surface modes, see \cite{barker15}.

%refer the reader to BL15 for surface modes

### 2015-03-28 paper.tex
updating due to barker latter 2015

In this work we study the linear VSI in radially local, vertically
global models of astrophysical disks. This approach has
been adopted in recent studies of the linear VSI
\citep{nelson13,mcnally14,barker15}. However, these studies consider 
isothermal perturbations, for which the thermal timescale is formally
zero, and/or neutrally stratified disks in which the stabilizing
effect of bouyancy is absent. We account for both effects to study the
VSI in more general disk models. To do so, we examine 
sub-adiabatically stratified disks and include an energy equation 
with a source term that allows the thermodynamic response of 
perturbations to be varied smoothly between isothermal and 
adiabatic.

>>>>

In this work we study the linear VSI in radially local, vertically
global models of astrophysical disks. This approach has
been adopted in recent studies of the linear VSI
\citep{nelson13,mcnally14,barker15}. However, these studies consider 
isothermal perturbations, for which the thermal timescale is formally
zero, and/or neutrally stratified disks in which the stabilizing
effect of buoyancy is absent. We account for both effects to study the
VSI in more general disk models. To do so, we examine 
sub-adiabatically stratified disks and include an energy equation 
with a source term that allows the thermodynamic response of 
perturbations to be varied smoothly between isothermal and 
adiabatic.
